\documentclass[dvips,aoas,preprint]{imsart}
%\documentclass[dvips,aoas]{imsart}

\RequirePackage[OT1]{fontenc}
\RequirePackage{amsthm,amsmath}
\RequirePackage[numbers]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

\usepackage{amssymb,amsbsy}
\usepackage{mathtools}  % for \coloneqq
%\usepackage[dvips]{graphics}
\usepackage{graphicx}
\usepackage{rotating}  % for sidewaystable

% Packages that aren't standard for AOAS:
%\usepackage{subfig} 

% Watermarking:
%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkLightness{0.9}
%\SetWatermarkScale{5}  % default is 1.2

% To typeset latex command names:
\newcommand\cmdname[1]{{\tt \textbackslash#1}}
\chardef\bslash=`\_

%\usepackage{amsthm,amsmath,natbib}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

% provide arXiv number if available:
%\arxiv{math.PR/0000000}

%===============================================================================
% put your definitions there:
\startlocaldefs

% Macro for an "editorial note"; uncomment the 2nd definition to
% produce a version without notes (put it in main doc if desired).




\newcommand\enote[1]{{$\bullet\bullet\bullet$}{\sl [#1]}{$\bullet\bullet\bullet$}}

%\renewcommand\enote[1]{\relax}





% Marginal note (just a redef of \marginpar or \marginparodd).
% Uncomment the 2nd definition to produce a version without notes (put it
% in main doc if desired).
% Could use \footnotesize for a bigger font.
% The \hspace enables hyphenation of the 1st word.
\setlength{\marginparwidth}{.85in}
\newcommand\mnote[1]{\-\marginpar[\raggedleft\scriptsize\hspace*{0pt}#1]%
{\raggedright\scriptsize\hspace*{0pt}#1}}
%\renewcommand\mnote[1]{\relax}

\newcommand{\txw}{\textwidth}

\renewcommand{\arraystretch}{1.2}


\newcommand{\bm}[1]{\mbox{\boldmath{$#1$}}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{eqnarray}}
\newcommand{\ea}{\end{eqnarray}}
\newcommand{\barr}{\begin{array}}
\newcommand{\earr}{\end{array}}
\newcommand{\bc}{\begin{center}}
\newcommand{\ec}{\end{center}}

\newcommand{\like}{{\cal L}}
\newcommand{\pval}{$p$-value}
\newcommand{\Eth}{E_{\rm th}}
\newcommand{\Dmax}{D_{\rm max}}
\newcommand{\Nsrc}{N_S}

% Coincidence model macros; t* = target, h* = host:
\newcommand{\tdrxn}{\omega}  % target (CR) direction
\newcommand{\hdrxn}{\varpi}  % host (AGN) direction
\newcommand{\expo}{\epsilon}  % exposure
\newcommand{\tlabel}{\lambda}  % target label
\newcommand{\Fvec}{\bm{F}}  % host fluxes
\newcommand{\Aperp}{A_\perp}  % perp. area

\newcommand{\D}{\overline{\mbox{D}}}

\newcommand{\simless}{\mathbin{\lower 3pt\hbox
   {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\newcommand{\simgreat}{\mathbin{\lower 3pt\hbox
   {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}}
\newcommand{\gta}{\simgreat}
\newcommand{\lta}{\simless}
\newcommand\msun{{\rm M}_\odot}
\endlocaldefs

%===============================================================================
\begin{document}

%===============================================================================
\begin{frontmatter}

% "Title of the paper"
\title{Multilevel Bayesian Framework for Modeling the Production,
Propagation and Detection of Ultra High Energy Cosmic Rays}

% For footnote indicating draft status:
% END OF TITLE TEXT\protect\thanksref{DRAFT}}
%\thankstext{DRAFT}{This is an description of work in progress, prepared for
%participants in the 2011 workshop, {\em Case Studies in Bayesian Statistics and
%Machine Learning}.  {\bf PLEASE DO NOT CIRCULATE!}}

% indicate corresponding author with \corref{}
% \author{\fnms{John} \snm{Smith}\corref{}\ead[label=e1]{smith@foo.com}\thanksref{t1}}
% \thankstext{t1}{Thanks to somebody}
% \address{line 1\\ line 2\\ printead{e1}}
% \affiliation{Some University}

\author{\fnms{Kunlaya} \snm{Soiaporn}\corref{}\ead[label=e1]{ks354@cornell.edu}},
\address{\printead{e1}}
\author{\fnms{Thomas} \snm{Loredo}\ead[label=e2]{loredo@astro.cornell.edu}},
\address{\printead{e2}}
\author{\fnms{David} \snm{Chernoff}\ead[label=e3]{cheroff@astro.cornell.edu}},
\address{\printead{e3}}
\author{\fnms{David} \snm{Ruppert}\ead[label=e4]{dr24@cornell.edu}}
\address{\printead{e4}}
\and
\author{\fnms{Ira} \snm{Wasserman}\ead[label=e5]{ira@astro.cornell.edu}}
\address{\printead{e5}}
\affiliation{Cornell University}


\runauthor{Soiaporn, et al.}
\runtitle{Multilevel Models for Cosmic Rays}

\begin{abstract}
Cosmic rays are atomic nuclei with velocities close to the speed of light.
Ultra-high energy cosmic rays (UHECRs) are approximately ten million times
more energetic than the most extreme particles produced at the Large Hadron
Collider and likely originate from nearby extragalactic sources. Important
astrophysical questions include: what phenomenon accelerates particles to such
large energies, which astronomical objects host the accelerators, and what
sorts of nuclei end up being energized? We develop a multilevel Bayesian
framework for assessing evidence for association of UHECRs and candidate source
populations that has four levels:
(1) a source property description level, including specification of a
candidate source population and the distribution of cosmic ray
intensities among sources;
(2) a cosmic ray production level, using a marked Poisson point process
to model latent properties (arrival times and energies) of cosmic rays;
(3) a propagation level, modeling deflection of cosmic ray trajectories by
cosmic magnetic fields;
(4) a detection and measurement level, accounting for detector efficiency
and exposure, and measurement errors in arrival directions and energies.
\mnote{Altered abstract from here}
Our framework can flexibly accomodate astrophysical components anywhere on
a spectrum from simplicity to great complexity and realism.
The number of possible associations is huge even for modest-sized datasets. 
We describe a Markov chain Monte Carlo algorithm for estimating model
parameters and comparing models by computing, via Chib's method, marginal
likelihoods and Bayes factors.

We demonstrate the framework by using simple model components to analyze
observations of 69 UHECRs by the Pierre Auger Observatory (PAO) during the
years 2004--2009, using a volume-complete catalog of 17 nearby active
galactic nuclei as a candidate host population. The reported data are
incomplete; an early portion of the data (``Period~1,'' with 14 events) was
used to set an energy cut maximizing a measure of anisotropy in that data;
only data for cosmic rays with energies above that cut are reported.  Also,
measurement errors are approximately summarized.  These factors are
problematic for independent analyses of PAO data.  Within the context of
``standard candle'' source models (all sources with the same isotropic
cosmic ray emission rate), and considering only the untuned data subsequent
to Period~1, there is no significant evidence favoring association of UHECRs
with nearby AGN vs.\ attributing them to sources drawn from an isotropic
background distribution.  The highest-probability associations are with the
two nearest AGN, Centaurus~A and its neighbor, NGC~4945.  If the association
model is adopted, the fraction of UHECRs that may be associated is likely
nonzero but is constrained to be well below 50\%.  Relatively modest magnetic
deflection angular scales of $\approx 3^\circ$ to $30^\circ$ are favored;
models that assign a large fraction of UHECRs to a single nearby source are
ruled out unless very large deflection scales are specified a priori, and
even then they are disfavored.  However, including the Period~1 data alters
the conclusions significantly, and a simulation study supports the idea that
the Period~1 data are anomalous, presumably due to the tuning.  Accurate and
optimal analysis of future data will likely require more complete disclosure
of the data.
\end{abstract}


%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\begin{keyword}
\kwd{Multilevel modeling}
\kwd{Hierarchical Bayes}
\kwd{Astrostatistics}
\kwd{Cosmic rays}
\kwd{Directional data}
\kwd{Coincidence assessment}
\kwd{Bayes factors}
\end{keyword}

\end{frontmatter}

%===============================================================================

%\enote{Editorial notes appear like this, using the \cmdname{enote} macro
%defined in paper.tex.  For issues not resolved but being deferred, leave the
%\cmdname{enote} in place, but uncomment the \cmdname{renewcommand}
%redefinition of \cmdname{enote} in paper.tex to hide them; we'll deal with
%them later.}


\input{intro-arxiv}

\input{data-arxiv}

\input{modeling-arxiv}

\input{results}

\input{checking-arxiv}

\input{discussion}

\appendix

\input{app-exposure}

\input{app-like}

\input{app-computation}

\input{app-CenA}

\input{app-comparison}

\input{app-chgpt-arxiv}

\section*{Acknowledgements}
This research was supported by the NSF via grants AST-0908439 and DMS-0805975.
We are grateful to Paul Sommers for helpful conversations about the
PAO instrumentation and data reduction and analysis process.

% Descriptions of online supplements go here:

%\begin{supplement}[id=suppA]
%  \sname{Supplement A}
%  \stitle{Title}
%  \slink[url]{http://lib.stat.cmu.edu/aoas/???/???}
%  \sdescription{Some text}
%\end{supplement}


\bibliographystyle{plain}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\jcap}{Journal of Cosmology and Astroparticle Physics}

\bibliography{UHECR}

% AOS,AOAS: If there are supplements please fill:
%\begin{supplement}[id=suppA]
%  \sname{Supplement A}
%  \stitle{Title}
%  \slink[url]{http://lib.stat.cmu.edu/aoas/???/???}
%  \sdescription{Some text}
%\end{supplement}

\end{document}

Chib, S. (1995) Marginal likelihood from the Gibbs Output, JASA, 90,
1313—1321.

Newton, M. A., and Raftery, A. E. (1994) Approximate Bayesian inference by the
weighted likelihood bootstrap, JRSS-B, 56, 3-48.

Cook, Gelman, Rubin 2006

