\documentclass[dvips,aoas,preprint]{imsart}
%\documentclass[dvips,aoas]{imsart}

\RequirePackage[OT1]{fontenc}
\RequirePackage{amsthm,amsmath}
\RequirePackage[numbers]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

\usepackage{amssymb,amsbsy}
\usepackage{mathtools}  % for \coloneqq
%\usepackage[dvips]{graphics}
\usepackage{graphicx}
\usepackage{rotating}  % for sidewaystable

% Packages that aren't standard for AOAS:
%\usepackage{subfig} 

% Watermarking:
%\usepackage{draftwatermark}
%\SetWatermarkText{DRAFT}
%\SetWatermarkLightness{0.9}
%\SetWatermarkScale{5}  % default is 1.2

% To typeset latex command names:
\newcommand\cmdname[1]{{\tt \textbackslash#1}}
\chardef\bslash=`\_

%\usepackage{amsthm,amsmath,natbib}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

% provide arXiv number if available:
%\arxiv{math.PR/0000000}

%===============================================================================
% put your definitions there:
\startlocaldefs

% Macro for an "editorial note"; uncomment the 2nd definition to
% produce a version without notes (put it in main doc if desired).




\newcommand\enote[1]{{$\bullet\bullet\bullet$}{\sl [#1]}{$\bullet\bullet\bullet$}}

%\renewcommand\enote[1]{\relax}





% Marginal note (just a redef of \marginpar or \marginparodd).
% Uncomment the 2nd definition to produce a version without notes (put it
% in main doc if desired).
% Could use \footnotesize for a bigger font.
% The \hspace enables hyphenation of the 1st word.
\setlength{\marginparwidth}{.85in}
\newcommand\mnote[1]{\-\marginpar[\raggedleft\scriptsize\hspace*{0pt}#1]%
{\raggedright\scriptsize\hspace*{0pt}#1}}
%\renewcommand\mnote[1]{\relax}

\newcommand{\txw}{\textwidth}

\renewcommand{\arraystretch}{1.2}


\newcommand{\bm}[1]{\mbox{\boldmath{$#1$}}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{eqnarray}}
\newcommand{\ea}{\end{eqnarray}}
\newcommand{\barr}{\begin{array}}
\newcommand{\earr}{\end{array}}
\newcommand{\bc}{\begin{center}}
\newcommand{\ec}{\end{center}}

\newcommand{\like}{{\cal L}}
\newcommand{\pval}{$p$-value}
\newcommand{\Eth}{E_{\rm th}}
\newcommand{\Dmax}{D_{\rm max}}
\newcommand{\Nsrc}{N_S}

% Coincidence model macros; t* = target, h* = host:
\newcommand{\tdrxn}{\omega}  % target (CR) direction
\newcommand{\hdrxn}{\varpi}  % host (AGN) direction
\newcommand{\expo}{\epsilon}  % exposure
\newcommand{\tlabel}{\lambda}  % target label
\newcommand{\Fvec}{\bm{F}}  % host fluxes

\newcommand{\D}{\overline{\mbox{D}}}

\newcommand{\simless}{\mathbin{\lower 3pt\hbox
   {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\newcommand{\simgreat}{\mathbin{\lower 3pt\hbox
   {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}}
\newcommand{\gta}{\simgreat}
\newcommand{\lta}{\simless}
\newcommand\msun{{\rm M}_\odot}
\endlocaldefs

%===============================================================================
\begin{document}

%===============================================================================
\begin{frontmatter}

% "Title of the paper"
\title{Multilevel Bayesian Framework for Modeling the Production,
Propagation and Detection of Ultra High Energy Cosmic Rays}

% For footnote indicating draft status:
% END OF TITLE TEXT\protect\thanksref{DRAFT}}
%\thankstext{DRAFT}{This is an description of work in progress, prepared for
%participants in the 2011 workshop, {\em Case Studies in Bayesian Statistics and
%Machine Learning}.  {\bf PLEASE DO NOT CIRCULATE!}}

% indicate corresponding author with \corref{}
% \author{\fnms{John} \snm{Smith}\corref{}\ead[label=e1]{smith@foo.com}\thanksref{t1}}
% \thankstext{t1}{Thanks to somebody}
% \address{line 1\\ line 2\\ printead{e1}}
% \affiliation{Some University}

\author{\fnms{Kunlaya} \snm{Soiaporn}\corref{}\ead[label=e1]{ks354@cornell.edu}},
\address{\printead{e1}}
\author{\fnms{Thomas} \snm{Loredo}\ead[label=e2]{loredo@astro.cornell.edu}},
\address{\printead{e2}}
\author{\fnms{David} \snm{Chernoff}\ead[label=e3]{cheroff@astro.cornell.edu}},
\address{\printead{e3}}
\author{\fnms{David} \snm{Ruppert}\ead[label=e4]{dr24@cornell.edu}}
\address{\printead{e4}}
\and
\author{\fnms{Ira} \snm{Wasserman}\ead[label=e5]{ira@astro.cornell.edu}}
\address{\printead{e5}}
\affiliation{Cornell University}


\runauthor{Soiaporn, et al.}
\runtitle{Multilevel Models for Cosmic Rays}

\begin{abstract}
Cosmic rays are atomic nuclei with velocities close to the speed of light.
Ultra-high energy cosmic rays (UHECRs) are approximately ten million times
more energetic than the most extreme particles produced at the Large Hadron
Collider and likely originate from nearby extragalactic sources. Important
astrophysical questions include: what phenomenon accelerates particles to such
large energies, which astronomical objects host the accelerators, and what
sorts of nuclei end up being energized? We develop a multilevel Bayesian
framework for assessing evidence for association of UHECRs and candidate source
populations that has four levels:
(1) a source property description level, including specification of a
candidate source population and the distribution of cosmic ray
intensities among sources;
(2) a cosmic ray production level, using a marked Poisson point process
to model latent properties (arrival times and energies) of cosmic rays;
(3) a propagation level, modeling deflection of cosmic ray trajectories by
cosmic magnetic fields;
(4) a detection and measurement level, accounting for detector efficiency
and exposure, and measurement errors in arrival directions and energies.
Our framework can flexibly accomodate astrophysical components anywhere on
a spectrum from simplicity to great complexity and realism.  
We demonstrate the framework using simple models similar to those
adopted in previous studies, but our results suggest more complex models
are likely necessary (we will pursue such models in future work).

The principal obstacle to computing with this framework is the
combinatorial explosion in the number of possible associations as the
sizes of the candidate source population and the UHECR sample grow.  We
present a Markov chain Monte Carlo algorithm for estimating the
parameters and computing, via Chib's method, the marginal likelihoods
used in model comparison.  We apply the algorithm to the observations of
69 UHECRs observed by Pierre Auger Observatory (PAO) during the years
2004--2009, using a volume-complete catalog of 17 nearby active galactic
nuclei as a candidate host population.
The reported data are incomplete; an early portion of the data
(``Period~1'') was used to set an energy cut maximizing a measure of
anisotropy in that data; only data for cosmic rays with energies above
that cut is reported, in Period~1 (14 events) and subsequently.  Also,
measurement errors are approximately summarized.  These factors are
problematic for independent analyses of PAO data.  Within the context of
``standard candle'' source models (all sources with the same isotropic
cosmic ray emission rate), there is no significant evidence favoring
association of UHECRs with nearby AGN vs.\ attributing them to sources
drawn from an isotropic background distribution, when we analyze the
untuned data subsequent to Period~1; the highest-probability
associations are with the two nearest AGN, Centaurus~A and its neighbor,
NGC~4945.  If the association model is adopted, the fraction of UHECRs
that may be associated is likely nonzero but is constrained to be well
below 50\%.  Relatively small magnetic deflection angular scales of
$\approx 3^\circ$ to $20^\circ$ are favored; models that assign a large
fraction of UHECRs to a single nearby source are ruled out unless very
large deflection scales are specified a priori, and even then they are
disfavored.  However, including the Period~1 data alters the conclusions
significantly, and a simulation study supports the idea that the
Period~1 data are anomalous, presumably due to the tuning.  Accurate and
optimal analysis of future data will likely require more complete
disclosure of the data.
\end{abstract}


%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\begin{keyword}
\kwd{Multilevel modeling}
\kwd{Hierarchical Bayes}
\kwd{Astrostatistics}
\kwd{Cosmic rays}
\kwd{Directional data}
\kwd{Coincidence assessment}
\kwd{Bayes factors}
\end{keyword}

\end{frontmatter}

%===============================================================================

%\enote{Editorial notes appear like this, using the \cmdname{enote} macro
%defined in paper.tex.  For issues not resolved but being deferred, leave the
%\cmdname{enote} in place, but uncomment the \cmdname{renewcommand}
%redefinition of \cmdname{enote} in paper.tex to hide them; we'll deal with
%them later.}


\input{intro-aoas}

\input{data-aoas}  % omits extra skymaps

\input{modeling}

\input{results}

% This is in place of the model checking section, which appears here in
% the arXiv version:


\subsection{Model checking}

\mnote{New subsection; model checking section moved to appendix.}
In Online Appendix~\ref{app:checking} we describe results of two types of
tests of our models, motivated by period-to-period variability of some of
the results reported above.

First, we performed simple change point analyses to see whether the
period-to-period variation of the Bayes factors for association vs.\ isotropy
indicates the population-level properties of the detected cosmic rays vary
from period to period.  We compared versions of $M_1$ and $M_2$ that allow
model parameters to change between periods to versions that keep the
parameters the same for all periods.  We find that there is no significant
evidence for variability of model parameters from period to period.  

%That is, presuming one of the models is adequate, the apparent discrepancy
%among the Bayes factors in Table~\ref{BFTable} reflects variability that may
%be expected for these modest sample sizes.

Second, we performed predictive checks to see whether the period-to-period
Bayes factor variations are surprising in the context of either the null or
association models, essentially using the Bayes factors as goodness-of-fit
test statistics.  
We simulated data from the null (isotropic) model and compared the Bayes
factors based on the observed data with those found in the simulations; we
did the same for a representative association model.
We find that Bayes factors favoring association as large as that
found with the Period~1 PAO data are unlikely for
isotropic models.
This implies the distribution of directions in the Period~1 sample is
anisotropic, but the calculation does not address whether this may be due to
tuning or to genuine anisotropy.
For association models, the large Bayes factors for periods 1 and 2,
and the small Bayes factor in Period~3, are not individually surprising.
But it is very surprising to see a combination of large Bayes factors
for the two small subsamples, and a small Bayes factor for the large
subsample.  The full dataset thus is not comfortably fit by either
isotropic or association models.  We discuss this further below.


\input{discussion}

\appendix

\input{app-exposure}

\input{app-like}

\input{app-computation}

\input{app-CenA}

\input{app-comparison}

\input{app-checking-aoas}


\section*{Acknowledgements}
This research was supported by the NSF via grants AST-0908439 and DMS-0805975.
We are grateful to Paul Sommers for helpful conversations about the
PAO instrumentation and data reduction and analysis process.

% Descriptions of online supplements go here:

%\begin{supplement}[id=suppA]
%  \sname{Supplement A}
%  \stitle{Title}
%  \slink[url]{http://lib.stat.cmu.edu/aoas/???/???}
%  \sdescription{Some text}
%\end{supplement}


\bibliographystyle{plain}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\jcap}{Journal of Cosmology and Astroparticle Physics}

\bibliography{UHECR}

% AOS,AOAS: If there are supplements please fill:
%\begin{supplement}[id=suppA]
%  \sname{Supplement A}
%  \stitle{Title}
%  \slink[url]{http://lib.stat.cmu.edu/aoas/???/???}
%  \sdescription{Some text}
%\end{supplement}

\end{document}

Chib, S. (1995) Marginal likelihood from the Gibbs Output, JASA, 90,
1313—1321.

Newton, M. A., and Raftery, A. E. (1994) Approximate Bayesian inference by the
weighted likelihood bootstrap, JRSS-B, 56, 3-48.

Cook, Gelman, Rubin 2006

